Epoch of Reionization Power Spectrum Results from PAPER-128

### 2016-07-26 PAPER.tex
first pass tidying of current draft. Put {rm} around units, {sc} around omnical as a software name and {tt} around stages of that software package, e.g. for FirstCal. Fixed spelling errors.

\documentclass[preprint2,numberedappendix,tighten,twocolappendix]{aastex6}  % USE THIS TO MAKE BIB, THEN FORMAT USING EMULATEAPJ
%\documentclass[twocolumn,apj,numberedappendix]{emulateapj}
\shorttitle{Epoch of Reionization Power Spectrum Results from PAPER-128}
\shortauthors{Cheng}

@@

\author{
%Aaron R. Parsons\altaffilmark{1,2},
%Adrian Liu\altaffilmark{1,2},
%James E. Aguirre\altaffilmark{3},
%Zaki S. Ali\altaffilmark{1},
Carina Cheng\altaffilmark{1}
%Daniel C. Jacobs\altaffilmark{8},
%David F. Moore\altaffilmark{3},
%Jonathan C. Pober\altaffilmark{1},
}

@@

The EoR has not yet been directly detected. However, constraints have been made on its timing and duration through different methods. As reionization progressed, the number of free electrons susceptible to Thompson scattering increased, resulting in scattered cosmic microwave background (CMB) photons whose imprint has been left on CMB observations. In particular, the Thompson scattering optical depth parameter $\tau$ is a useful metric for characterizing the duration of reionization. Recent Planck measurements of $\tau$ suggest a late reionization history, occurring between redshifts of $7.7$ and $8.7$.

Galaxy observations have also begun to place constraints on the EoR. With their high luminosities, quasars make for powerful probes of the state of the IGM through absorption line studies. As photons from a quasar are redshifted, they may be absorbed by neutral hydrogen at the Lyman-$\alpha$ line, producing Gunn-Peterson trough features in quasar spectra. Observations of these spectra in high redshift galaxies have signaled the end of reionization by $z ~ 6$, and experiments are continuously pushing galaxy observations to higher redshifts.

A third powerful probe of the EoR makes use of the $21 cm$ wavelength emission produced by hydrogen due to its two quantum states. Because hydrogen is so pervasive in the Universe and a transition from its spin up state to spin down state gives off a  well-defined energy difference in the form of a $21 cm$ wavelength photon, observations targeting this signal have the potential to trace the EoR from start to finish. However, $21 cm$ experiments face major challenges, including those from bright foregrounds and instrumental systematics. Nevertheless, the observations and analyses of data from current radio telescope arrays have already begun to constrain the $21 cm$ signal and make statements about the nature of the IGM during the EoR.

Major existing $21 cm$ experiments include both those targeting to measure the sky-averaged `global' $21 cm$ signal and those employing interferometric elements and the use of cross-correlations and statistical power-spectral measurements. Examples of the former are EDGES, the LWA, LEDA, DARE, BigHorns, and SARAS. Major interferometers include the GMRT, LOFAR, the MWA, the 21CMA, and PAPER.

In this paper we present results from the $128$ antenna configuration of the Donald C. Baker Precision Array for Probing the Epoch of Reionization (PAPER). PAPER is a highly redundant array consisting of dual-polarization dipole antennas. Its configuration is optimal for boosting power spectrum sensitivity and the analysis used relies on a foreground avoidance approach where we measure power spectrum values in a region of delay space that is uncontaminated by foregrounds. Previous upper limits on the EoR power spectrum include those made by PAPER-32, which used its redundant configuration to its advantage to obtain an upper limit of $\Delta^{2}(k) \leq (41 mk)^{2}$ at $k=0.28h Mpc^{-2}$. Doubling in array size and employing an updated analysis pipeline that included fringe-rate filtering and an optimal quadratic estimator framework for the power spectrum analysis achieved an upper limit on $\Delta^{2}(k)$ of $(22.4 mK)^{2}$ in the range $0.15 < k < 0.5h Mpc^{-1}$ at $z = 8.4$. The result from PAPER-64 was the best published upper limit thus far, with PAPER-128 achieving a factor of ?? in improvement.

@@

\begin{figure*}[!]
	\centering
	\includegraphics[width=1.0\textwidth]{antlayout.png}
	\caption{The antenna layout of PAPER-128. Core grid antennas are labeled $1$ through $111$ and outrigger antennas are labeled $112$ through $127$.}
	\label{fig:layout}
\end{figure*}

@@

Raw PAPER data consists of $1024$ frequency channels sampled every $97.6 kHz$. Every $10$ minute file then goes through our compression pipeline, which consists of a two-step radio frequency interference (RFI) flagger and averaging along both the frequency (delay filtering) and time (fringe rate filtering) axes. Final compressed data consists of $203$ frequency channels covering a bandwidth of $100-200 MHz$. Each $10$ minute file contains data sampled at integrations of $31.65 s$. The amplitudes of one night of raw and compressed data is shown in \ref{fig:raw} as waterfall plots of frequency versus time.

\begin{figure*}[!]
	\centering
	%\includegraphics[width=1.0\textwidth]{raw.png}
	\caption{Waterfall plots (in log Janskys) of one night of raw PAPER-128 data (left) and the same data after compression (right). Compression downsizes files by averaging over time and frequency and flags radio frequency interference.}
	\label{fig:raw}
\end{figure*}

@@

The sought-after $21 cm$ signal lies beneath bright foregrounds and various instrumental systematics that contaminate our data. Because foregrounds are predicted to be $4-5$ orders of magnitude brighter than the cosmological signal, $21 cm$ experiments require high dynamic ranges and considerable sensitivities. PAPER is designed to boost its sensitivity by containing many short, repeated baselines of the same length and orientation in its array. Thus, PAPER is able to measure the same Fourier-mode on the sky multiple times while beating down the amount of noise with each successive measurement.

@@

For our analysis, we use the package Omnical for redundant calibration. Similarly to the analysis of PAPER-64, Omnical comprises of three steps. The first is $FirstCal$, which uses all baseline redundancies to generate a static gain solution for each antenna that will unwrap any phase wrapping between two identical baselines. We perform $FirstCal$ because the next stage of Omnical cannot tell the difference between a phase of $0$ and $2\pi$, for example. The second step is $LogCal$, which takes the log of all the visibility equations and separates the real and imaginary components into two matrices. Coarse solutions are determined for both the antenna gains and `model' visibilities (one for each baseline type) simultaneously. The final step of Omnical is $LinCal$, which applies small perturbations to the $LogCal$ solutions in an iterative fashion, honing in on the optimal solutions.

We perform redundant calibration for each epoch of data used in our analysis, as each epoch has to be separately calibrated since they are on different overall gain scales. We divide out the antenna gains for each of our measured visibilities, bringing them into alignment with each other (for identical baseline types) and the `model' visibility for that baseline type. We also include the subtraction of crosstalk in our dataset. Crosstalk shows up as a constant signal in time in our visibilities and is attributed to signals from one antenna being picked up by another. Omnical handles crosstalk by subtracting off an averaged residual (differences between measured visibilities and the `model' visibility, averaged over baseline type) for each visibility measurement.

For PAPER-128, we update the Omnical software package, originally written in C, with a user-friendly interface written in Python.  Additionally, we employ a chi-square metric on a per antenna basis in order to exclude dead or faulty antennas and improve our calibration solutions. \cc{Add plots to show stability of Omnical solutions and chi-square examples}

It is important to note that while Omnical is powerful for ensuring array redundancy, it is not able to solve for $4$ calibration parameters - namely, to correct for the overall gain, phase, and tip/tilt of the array. In order to do this, we turn to...

@@

Foregrounds pose a significant challenge to EoR experiments. Some experiments, like LOFAR and the MWA, rely on modeling and foreground subtraction to create a clean EoR `window' in which to make their power spectrum measurements. Previous PAPER results, including this one, instead focus on a foreground avoidance approach.

In order to filter foregrounds, we exploit their spectrally-smooth nature. By Fourier-transforming our visibilities along the frequency axis, the delay domain provides a natural space in which to isolate foregrounds. Physically, a time delay can be thought of as the extra time light from a source takes to reach one antenna compared to another one (due to slightly different traveling distances). The maximum delay a source on the sky can have is if it is located on the horizon. In this case, the delay is determined by the length of the baseline. Consequently, all spectrally-smooth foreground signals show up within a maximum delay (or `horizon limit') given a baseline length. The EoR signal, however, is believed to mimic noise which has a random frequency structure and is not confined to be within the horizon limit. A simple way to avoid foregrounds without removing the cosmological signal, therefore, is to make our power spectrum measurements outside of the horizon limit that contains the bright foregrounds.

In practice, we filter our calibrated visibilities by de-convolving out our sampling function (which contains flags due to RFI) from our delay-domain visibilities using a CLEAN-like algorithm and restricting our clean components to inside the horizon limit (set by the baseline length) plus a $15 ns$ buffer. The Fourier-transformed clean components are then subtracted from our visibilities. This filtering process is performed on a per-baseline, per-integration basis, and we achieve a brightness suppression of $~2$ orders of magnitude in our visibilities.

@@

Recalling that redundant calibration through Omnical does not solve for the overall calibration parameters of our array, we calibrate these final unknowns by matching our LST-binned datasets with that of PAPER-64. For the PAPER-64 analysis, absolute calibration was carried out by standard radio astronomy self-calibration. Overall phase terms were fitted using Pictor A, Fornax A, and the Crab Nebula, and the overall flux scale was determined using Pictor A as a calibrator source. PAPER-64's analysis produced even and odd LST-binned data covering an LST range of $0-8$, fully encompassing the LST ranges we use in our analysis.

@@

With our final visibilities in hand, we are ready to estimate the $21 cm$ power spectrum, $\hat{\textbf{p}}(\textbf{k})$. The power spectrum measurement is a function of $\textbf{k} = (k_{x},k_{y},k_{z})$, where each component of $k$ represents the Fourier-dual to spatial coordinates $(x,y,z)$. Our visibilities have already been Fourier-transformed along two of these axes, namely, along $k_{\perp} = \sqrt{k_{x}^{2}+k_{y}^{2}}$. This perpendicular component of $k$ is analogous to the spatial Fourier-modes in the plane of the sky, while the parallel component of $k$, $k_{\parallel}$, is the Fourier-transform of our visibilities along the line-of-sight (in the frequency direction).

Our visibilities measure the sky temperature in units of Janskys. Recalling that we are interested in the power of the cosmological signal on different $k$-scales, we can map our data to a power spectral measurement using the standard definition

@@

\noindent where $\tilde{T}_{b}(\textbf{k})$ is the spatial Fourier-transform of the brightness temperature of the sky, $\delta^{D}$ is the Dirac delta function, and $\langle \rangle$ is an ensemble average. As a simplistic explanation, we need to take our visibility temperature measurements, Fourier-transform them along the frequency axis and convert from units of Janskys to Kelvins, square, and bin in constant contours of $k$. In practice, however, we use linear algebra techniques as explained in the following section in order to obtain both $\hat{\textbf{p}}(\textbf{k})$ and its associated error bars.

@@

Both the fringe-rate filtering and inverse covariance weighting of our visibilities produces data that is susceptible to signal loss. This is due to the reduction of independent time samples from fringe-rate filtering which leads to a more noisy dataset. Consequently, when computing $\textbf{C}^{-1}$ from the data, the result from the inverse covariance weighting of that same data is to overfit the noise. This can produce power spectra values below the thermal noise limit that is predicted based on observation parameters.

@@

The ratio of $P_{in}$ to $P_{out}$ changes as the level of the injected signal \textbf{e} is increased. As shown in \cc{reference figure here}, when the injected signal level is low compared to the data \textbf{x}, the recovered output power spectrum is the same as the input. However, when \textbf{e} begins to dominate over  \textbf{x}, we move to a regime where the input power spectrum is greater than the output, implying signal loss due to the overfitting of noise during inverse covariance weighting.

@@

Our final $21 cm$ power spectrum measurements from PAPER-128 are for a redshift of $z=\cc{??}$. The data used for this result incorporates \cc{??} days and \cc{??} channels, centered around \cc{??}. We present both the power spectrum points from $\hat{\textbf{p}}(\textbf{k})$ in units of $mk^{2}(h^{-1}Mpc)^{3}$ and the corresponding dimensionless power spectrum points, $\Delta^{2}(\textbf{k})$, which is related to $\hat{\textbf{p}}(\textbf{k})$ by

>>>>

\documentclass[preprint2,numberedappendix,tighten,twocolappendix]{aastex6}  % USE THIS TO MAKE BIB, THEN FORMAT USING EMULATEAPJ
%\documentclass[twocolumn,apj,numberedappendix]{emulateapj}
\shorttitle{Epoch of Reionization Power Spectrum Results from PAPER-128}
\shortauthors{Cheng et al.}

@@

\author{
Carina Cheng\altaffilmark{1},
et al.
}

@@

The EoR has not yet been directly detected. However, constraints have been made on its timing and duration through different methods. As reionization progressed, the number of free electrons susceptible to Thompson scattering increased, resulting in scattered cosmic microwave background (CMB) photons whose imprint has been left on CMB observations. In particular, the Thompson scattering optical depth parameter $\tau$ is a useful metric for characterizing the duration of reionization. Recent \textit{Planck} measurements of $\tau$ suggest a late reionization history, occurring between redshifts of $7.7$ and $8.7$.

Galaxy observations have also begun to place constraints on the EoR. With their high luminosities, quasars make for powerful probes of the state of the IGM through absorption line studies. As photons from a quasar are redshifted, they may be absorbed by neutral hydrogen at the Lyman-$\alpha$ line, producing Gunn-Peterson trough features in quasar spectra. Observations of these spectra in high redshift galaxies have signalled the end of reionization by $z ~ 6$, and experiments are continuously pushing galaxy observations to higher redshifts.

A third powerful probe of the EoR makes use of the 21\,cm wavelength emission produced by hydrogen due to its two quantum states. Because hydrogen is so pervasive in the Universe and a transition from its spin up state to spin down state gives off a  well-defined energy difference in the form of a 21\,cm wavelength photon, observations targeting this signal have the potential to trace the EoR from start to finish. However, 21\,cm experiments face major challenges, including those from bright foregrounds and instrumental systematics. Nevertheless, the observations and analyses of data from current radio telescope arrays have already begun to constrain the 21\,cm signal and make statements about the nature of the IGM during the EoR.

Major existing 21\,cm experiments include both those targeting to measure the sky-averaged `global' 21\,cm signal and those employing interferometric elements and the use of cross-correlations and statistical power-spectral measurements. Examples of the former are EDGES, the LWA, LEDA, DARE, BIGHORNS, and SARAS. Major interferometers include the GMRT, LOFAR, the MWA, the 21CMA, and PAPER.

In this paper we present results from the $128$ antenna configuration of the Donald C. Baker Precision Array for Probing the Epoch of Reionization (PAPER). PAPER is a highly redundant array consisting of dual-polarization dipole antennas. Its configuration is optimal for boosting power spectrum sensitivity and the analysis used relies on a foreground avoidance approach where we measure power spectrum values in a region of delay space that is uncontaminated by foregrounds. Previous upper limits on the EoR power spectrum include those made by PAPER-32, which used its redundant configuration to its advantage to obtain an upper limit of $\Delta^{2}(k) \leq (41 {\rm mK})^{2}$ at $k=0.28h {\rm Mpc}^{-2}$. Doubling in array size and employing an updated analysis pipeline that included fringe-rate filtering and an optimal quadratic estimator framework for the power spectrum analysis achieved an upper limit on $\Delta^{2}(k)$ of $(22.4 {\rm mK}))^{2}$ in the range $0.15 < k < 0.5h {\rm Mpc}^{-1}$ at $z = 8.4$. The result from PAPER-64 was the best published upper limit thus far, with PAPER-128 achieving a factor of ?? in improvement.

@@

\begin{figure*}[!]
	\centering
	\includegraphics[width=1.0\textwidth]{antlayout.png}
	\caption{The antenna layout of PAPER-128. Core grid antennas are labeled $1$ through $111$ and outrigger antennas are labelled $112$ through $127$.}
	\label{fig:layout}
\end{figure*}

@@

Raw PAPER data consists of $1024$ frequency channels sampled every $97.6 {\rm kHz}$. Every $10$ minute file then goes through our compression pipeline, which consists of a two-step radio frequency interference (RFI) flagger and averaging along both the frequency (delay filtering) and time (fringe rate filtering) axes. Final compressed data consists of $203$ frequency channels covering a bandwidth of $100-200 {\rm MHz}$. Each $10$ minute file contains data sampled at integrations of $31.65 {\rm s}$. The amplitudes of one night of raw and compressed data is shown in \ref{fig:raw} as waterfall plots of frequency versus time.

\begin{figure*}[!]
	\centering
	%\includegraphics[width=1.0\textwidth]{raw.png}
	\caption{Waterfall plots (in log Jy) of one night of raw PAPER-128 data (left) and the same data after compression (right). Compression downsizes files by averaging over time and frequency and flags radio frequency interference.}
	\label{fig:raw}
\end{figure*}

@@

The sought-after 21\,cm signal lies beneath bright foregrounds and various instrumental systematics that contaminate our data. Because foregrounds are predicted to be $4-5$ orders of magnitude brighter than the cosmological signal, 21\,cm experiments require high dynamic ranges and considerable sensitivities. PAPER is designed to boost its sensitivity by containing many short, repeated baselines of the same length and orientation in its array. Thus, PAPER is able to measure the same Fourier-mode on the sky multiple times while beating down the amount of noise with each successive measurement.

@@

For our analysis, we use the package {\sc Omnical} for redundant calibration. Similarly to the analysis of PAPER-64, {\sc Omnical} comprises of three steps. The first is {\tt FirstCal}, which uses all baseline redundancies to generate a static gain solution for each antenna that will unwrap any phase wrapping between two identical baselines. We perform {\tt FirstCal} because the next stage of {\sc Omnical} cannot tell the difference between a phase of $0$ and $2\pi$, for example. The second step is {\tt LogCal}, which takes the log of all the visibility equations and separates the real and imaginary components into two matrices. Coarse solutions are determined for both the antenna gains and `model' visibilities (one for each baseline type) simultaneously. The final step of {\sc Omnical} is {\tt LinCal},, which applies small perturbations to the {\tt LogCal}, solutions in an iterative fashion, honing in on the optimal solutions.

We perform redundant calibration for each epoch of data used in our analysis, as each epoch has to be separately calibrated since they are on different overall gain scales. We divide out the antenna gains for each of our measured visibilities, bringing them into alignment with each other (for identical baseline types) and the `model' visibility for that baseline type. We also include the subtraction of crosstalk in our dataset. Crosstalk shows up as a constant signal in time in our visibilities and is attributed to signals from one antenna being picked up by another. {\sc Omnical} handles crosstalk by subtracting off an averaged residual (differences between measured visibilities and the `model' visibility, averaged over baseline type) for each visibility measurement.

For PAPER-128, we update the {\sc Omnical} software package, originally written in C, with a user-friendly interface written in Python.  Additionally, we employ a chi-square metric on a per antenna basis in order to exclude dead or faulty antennas and improve our calibration solutions. \cc{Add plots to show stability of {\sc Omnical} solutions and chi-square examples}

It is important to note that while {\sc Omnical} is powerful for ensuring array redundancy, it is not able to solve for $4$ calibration parameters - namely, to correct for the overall gain, phase, and tip/tilt of the array. In order to do this, we turn to...

@@

Foregrounds pose a significant challenge to EoR experiments. Some experiments, like LOFAR and the MWA, rely on modelling and foreground subtraction to create a clean EoR `window' in which to make their power spectrum measurements. Previous PAPER results, including this one, instead focus on a foreground avoidance approach.

In order to filter foregrounds, we exploit their spectrally-smooth nature. By Fourier-transforming our visibilities along the frequency axis, the delay domain provides a natural space in which to isolate foregrounds. Physically, a time delay can be thought of as the extra time light from a source takes to reach one antenna compared to another one (due to slightly different travelling distances). The maximum delay a source on the sky can have is if it is located on the horizon. In this case, the delay is determined by the length of the baseline. Consequently, all spectrally-smooth foreground signals show up within a maximum delay (or `horizon limit') given a baseline length. The EoR signal, however, is believed to mimic noise which has a random frequency structure and is not confined to be within the horizon limit. A simple way to avoid foregrounds without removing the cosmological signal, therefore, is to make our power spectrum measurements outside of the horizon limit that contains the bright foregrounds.

In practice, we filter our calibrated visibilities by de-convolving out our sampling function (which contains flags due to RFI) from our delay-domain visibilities using a CLEAN-like algorithm and restricting our clean components to inside the horizon limit (set by the baseline length) plus a $15 {\rm ns}$ buffer. The Fourier-transformed clean components are then subtracted from our visibilities. This filtering process is performed on a per-baseline, per-integration basis, and we achieve a brightness suppression of $~2$ orders of magnitude in our visibilities.

@@

Recalling that redundant calibration through {\sc Omnical} does not solve for the overall calibration parameters of our array, we calibrate these final unknowns by matching our LST-binned datasets with that of PAPER-64. For the PAPER-64 analysis, absolute calibration was carried out by standard radio astronomy self-calibration. Overall phase terms were fitted using Pictor A, Fornax A, and the Crab Nebula, and the overall flux scale was determined using Pictor A as a calibrator source. PAPER-64's analysis produced even and odd LST-binned data covering an LST range of $0-8$, fully encompassing the LST ranges we use in our analysis.

@@

With our final visibilities in hand, we are ready to estimate the 21\,cm power spectrum, $\hat{\textbf{p}}(\textbf{k})$. The power spectrum measurement is a function of $\textbf{k} = (k_{x},k_{y},k_{z})$, where each component of $k$ represents the Fourier-dual to spatial coordinates $(x,y,z)$. Our visibilities have already been Fourier-transformed along two of these axes, namely, along $k_{\perp} = \sqrt{k_{x}^{2}+k_{y}^{2}}$. This perpendicular component of $k$ is analogous to the spatial Fourier-modes in the plane of the sky, while the parallel component of $k$, $k_{\parallel}$, is the Fourier-transform of our visibilities along the line-of-sight (in the frequency direction).

Our visibilities measure the sky temperature in units of Jys. Recalling that we are interested in the power of the cosmological signal on different $k$-scales, we can map our data to a power spectral measurement using the standard definition

@@

\noindent where $\tilde{T}_{b}(\textbf{k})$ is the spatial Fourier-transform of the brightness temperature of the sky, $\delta^{D}$ is the Dirac delta function, and $\langle \rangle$ is an ensemble average. As a simplistic explanation, we need to take our visibility temperature measurements, Fourier-transform them along the frequency axis and convert from units of Jys to Kelvin, square, and bin in constant contours of $k$. In practice, however, we use linear algebra techniques as explained in the following section in order to obtain both $\hat{\textbf{p}}(\textbf{k})$ and its associated error bars.

@@

Both the fringe-rate filtering and inverse covariance weighting of our visibilities produces data that is susceptible to signal loss. This is due to the reduction of independent time samples from fringe-rate filtering which leads to a more noisy dataset. Consequently, when computing $\textbf{C}^{-1}$ from the data, the result from the inverse covariance weighting of that same data is to over-fit the noise. This can produce power spectra values below the thermal noise limit that is predicted based on observation parameters.

@@

The ratio of $P_{in}$ to $P_{out}$ changes as the level of the injected signal \textbf{e} is increased. As shown in \cc{reference figure here}, when the injected signal level is low compared to the data \textbf{x}, the recovered output power spectrum is the same as the input. However, when \textbf{e} begins to dominate over  \textbf{x}, we move to a regime where the input power spectrum is greater than the output, implying signal loss due to the over-fitting of noise during inverse covariance weighting.

@@

Our final 21\,cm power spectrum measurements from PAPER-128 are for a redshift of $z=\cc{??}$. The data used for this result incorporates \cc{??} days and \cc{??} channels, centered around \cc{??}. We present both the power spectrum points from $\hat{\textbf{p}}(\textbf{k})$ in units of $mk^{2}(h^{-1}Mpc)^{3}$ and the corresponding dimensionless power spectrum points, $\Delta^{2}(\textbf{k})$, which is related to $\hat{\textbf{p}}(\textbf{k})$ by
